Formalizing Stack Safety as a Security Property

### 2021-07-08 submission3/main.tex
pass through 6

The micro-policy we present here to illustrate enforcement of stack
safety as defined in \cref{sec:lse} is a small variation
on the {\em Depth Isolation} policy presented by
\citet{DBLP:conf/sp/RoesslerD18}. In their policy,  the stack memory is tagged
with ownership information \apt{based on depth!} associated to each stack frame (and to each
separate object inside that frame, a more finely grained access
control that we do not need to consider here), and tags registers containing
stack pointers with access permission information. The policy also
uses tags on instructions to identify the code
sequences that have permission to manage the stack, say during calls and returns.
%\leo{If we have the simple exposition so far, no need to mention the following sentence:}
%In the following, for ease of exposition, we will assume that no
%arguments are passed on the stack.
%% \rb{Conceivably on alloc
%%   operations, etc., which we have not yet mentioned.}

The micro-policy maintains tags of the form $\tagStackDepth{n}$ for
each stack location belonging to the stack frame at activation depth
$n$, or the tag $\tagNoDepth$ for stack locations that are currently
unused (but could potentially be used in the future).
%
The $\PCname$ is tagged similarly as $\tagPCDepth{n}$.
%
Initially, the entire stack is considered unused (and therefore tagged
$\tagNoDepth$), and the {\PCname} has tag $\tagPCDepth{0}$.
%
During normal execution, the micro-policy rules only permit load and
store operations when the target memory is tagged {\em with the same
  depth} as the current {\PCname} tag, and store operations also when the
target memory is $\tagNoDepth$.

More interesting changes to the tags (from a stack safety perspective)
happen at exactly the points in the program where we switch from a
caller to a callee or vice-versa:

@@

\item From caller to callee, when the machine executes an instruction marked as a call in
  the call map.
  At this point, the current
  $\tagPCDepth{n}$ tag is incremented to $\tagPCDepth{(n + 1)}$, and the function
  entry sequence initializes  all locations of the new function frame with
  tag $\tagStackDepth{(n + 1)}$.

\item From callee back to caller, when execution is about to reach a
  return point.
  %The exit sequence retags all locations in the function frame with $\tagNoDepth$.
  At the actual return instruction,
  the $\PCname$ tag is decremented.
%  \leo{Not sure the actual policy
%    does that now (retag unused). But I don't think it's a
%    confidentiality or integrity violation.}
%  \rb{Correct, and it doesn't. Some versions of the policy do clear
%    the stack, retagging would be easily added.}

@@

This discipline suffices to enforce local stack encapsulation.
To implement it,  we define blessed instruction sequences
intended to appear at the entry and exit of each function,
which manipulate tags as just described in addition to performing the
usual calling convention tasks of saving/restoring the return address to/from
the stack and adjusting the stack pointer. With the aid of an additional tag on
the $\rsp$ register, these sequences also serve to enforce
well-bracketed control flow.
The micro-policy guarantees atomic execution of these sequences
using a combination of tags on the instructions
and an additional tag on the $\PCname$.

There remains the question of how to ensure that the sequences are
invoked at the right places from the perspective of the stack-safety
property. For a program to enjoy stack safety
(with respect to a particular call map), all we need require is that
the entry sequence be initiated at any instruction marked as a call
in the call map. This is achieved by giving these instructions (another) special tag,
and it is easy to check statically that this has been done correctly.
If the code fails to initiate an exit
sequence at a point where the stack-safety property expects a return,
the micro-policy will incorrectly behave as if execution
is continuing in the callee, but since the callee never has stronger access
rights than the caller, this is harmless.

@@

We validate the micro-policy of the previous section by
systematically testing that it satisfies eager stack integrity and
confidentiality. We use a Coq specification of the RISC-V
architecture\footnote{Available at \url{https://github.com/mit-plv/riscv-coq}.}
%\leo{translated from Haskell via hs-2-coq?\apt{not worth saying since the MIT people did it}}
and extend it with a
runtime monitor implementing the stack-safety
micro-policy.\footnote{This code is {\em redacted for review}.\bcp{Why??}} We
  chose the Coq proof assistant as the setting for our implementation
  for two reasons: first, to ensure that our coinductive trace
  definitions are well-formed, and reason about them, and to leverage the power of the
  QuickChick property-based testing framework~\citep{Pierce:SF4}.

To use QuickChick, we build random test-case generators that produce
% \begin{enumerate}
% \item
(1) an initial RISC-V machine state, including most notably
  the program to be executed;
  % \item
(2)
  an initial policy state, tagging instructions corresponding to
  blessed call or return sequences appropriately, while marking
  all potential stack locations as $\tagNoDepth$; and
% \item
(3)
 a code map and a stack map, accurately characterizing the behavior
  of callers and callees.
% \end{enumerate}
%
That is, our generators need to produce not only a low-level program
that exhibits interesting control-flow patterns, but also the policy
meta-information and code annotations that would normally be produced
by the compiler.

@@

To write such generators we build on the work of
\citep{TestingNI:ICFP, DBLP:journals/jfp/HritcuLSADHPV16}, which
introduced {\em generation by execution} to produce progams leading
abstract machines towards interesting behaviors. Generation by
execution receives as an input a partially instantiated machine state
and attempts to generate an instruction (or a sequence of instructions
such as a blessed sequence) that makes sense locally (e.g., jumps go
to a potentially valid code location and loads read from from a
potentially valid stack location). Then we step the machine and repeat
the process until we generate or execute some target number of
instructions, or we reach a point where the machine can't step
anymore.

We extend this technique to keep track of the control flow behavior of
the program being generated: each time a call or return sequence is
generated, we ensure that the appropriate policy tags and code
annotations are set for the entry or return points. At the same time,
we allow the generation to sometimes relax those constraints,
introducing potentially ill-formed flows: this causes our programs
to failstop when executed in conjunction with the policy monitor,
but also allows for revealing errors in our setup.

@@

%The mean-time-to-failure (MTTF) for various bugs can be found in
%Table~\ref{tab:bug-table}. All experiments were run in a desktop
%machine equipped with i7-8700 CPU @ 3.20GHz with 32GB
%RAM.
% \leo{Something like: There is really no baseline to compare this
  %against, but include it here for the future/reasons?}
%
The artificially injected bugs weaken the tag propagation
and checking mechanism.
%
For example, when loading from a stack location, the correct
micro-policy needs to enforce that the tag on the location being read
is $\tagStackDepth{n}$ for some number $n$ and that the one of the
current $\PCname$ is $\tagPCDepth{n}$ for the same depth. We can relax
that restriction by not checking the depth equality (row {\em
  LOAD\_NO\_CHECK}).
%
Similarly, when storing to a stack location, the correct micro-policy
needs to ensure that the tag on the memory location is either
$\tagNoDepth$ or has again the same depth as the current $\PCname$
tag. Relaxing that constraint causes violations to the integrity
property (row {\em STORE\_NO\_CHECK}).

@@

\subsection{Why not Proof?}

There are several ways to evaluate whether an enforcement mechanism enforces
our properties. Ideally such validation would be done through formal proof.
Our model is certainly amenable to such a proof, with no fundamental barriers,
but it would be considerable work to carry out for a full ISA like RISC-V.
We concentrate instead on testing because it is better aligned with our immediate
goal of evaluating real enforcement mechanisms for real machine architectures.
\sna{More to say here?}

>>>>

The micro-policy we present here to illustrate enforcement of stack
safety as defined in \cref{sec:lse} is a small variation
on the {\em Depth Isolation} policy presented by
\citet{DBLP:conf/sp/RoesslerD18}. In their policies,  the stack memory is tagged
with ownership information associated to each stack frame (and to each
separate object inside that frame, a more finely grained access
control that we do not need to consider here), and tags registers containing
stack pointers with access permission information. The policy also
uses tags on instructions to identify the code
sequences that have permission to manage the stack, say during calls and returns.
%\leo{If we have the simple exposition so far, no need to mention the following sentence:}
%In the following, for ease of exposition, we will assume that no
%arguments are passed on the stack.
%% \rb{Conceivably on alloc
%%   operations, etc., which we have not yet mentioned.}

The Depth Isolation micro-policy maintains tags of the form $\tagStackDepth{n}$ for
each stack location belonging to the stack frame at activation depth
$n$, or the tag $\tagNoDepth$ for stack locations that are currently
unused (but could potentially be used in the future).
%
The $\PCname$ is tagged similarly as $\tagPCDepth{n}$.
%
Initially, the entire stack is considered unused (and therefore tagged
$\tagNoDepth$), and the {\PCname} has tag $\tagPCDepth{0}$.
%
During normal execution, the micro-policy rules only permit load and
store operations when the target memory is tagged {\em with the same
  depth} as the current {\PCname} tag, and store operations also when the
target memory is $\tagNoDepth$.

Interesting changes to the tags occur where the program switches from 
a caller to a callee or vice-versa:

@@

\item From caller to callee: when the machine executes a call,
  %, when the machine executes an instruction marked as a call in
  % the call map.
  the current
  $\tagPCDepth{n}$ tag is incremented to $\tagPCDepth{(n + 1)}$, and the function
  entry sequence initializes  all locations of the new function frame with
  tag $\tagStackDepth{(n + 1)}$.

\item From callee back to caller: %when execution is about to reach a
%  return point.
  %The exit sequence retags all locations in the function frame with $\tagNoDepth$.
  at the actual return instruction,
  the $\PCname$ tag is decremented. \apt{Doesn't the exit sequence retag to $\tagNoDepth$?}
%  \leo{Not sure the actual policy
%    does that now (retag unused). But I don't think it's a
%    confidentiality or integrity violation.}
%  \rb{Correct, and it doesn't. Some versions of the policy do clear
%    the stack, retagging would be easily added.}

@@

This discipline suffices to enforce stack safety. 
To implement it,  we define blessed instruction sequences
intended to appear at the entry and exit of each function,
which manipulate tags as just described in addition to performing the
usual calling convention tasks of saving/restoring the return address to/from
the stack and adjusting the stack pointer. With the aid of an additional tag on
the $\rsp$ register, these sequences also serve to enforce
well-bracketed control flow.
The micro-policy guarantees atomic execution of these sequences
using a combination of tags on the instructions
and an additional tag on the $\PCname$.

There remains the question of how to ensure that the sequences are
invoked at the right places from the perspective of the stack-safety
property. For a program to enjoy stack safety
(with respect to a particular call map), all we need require is that
the entry sequence be initiated at \apt{immediately following?}  any instruction marked as a call
in the call map. This is achieved by giving these instructions (another) special tag,
and it is easy to check statically that this has been done correctly.
If the code fails to initiate an exit
sequence at a point where the stack-safety property expects a return,
the micro-policy will incorrectly behave as if execution
is continuing in the callee, but since the callee never has stronger access
rights than the caller, this is harmless.

@@

There are several ways to evaluate whether an enforcement mechanism enforces
stack safety properties. Ideally such validation would be done through formal proof over
the semantics of the enforcement-augmented machine. 
However, while there are no fundamental barriers to producing such a proof, 
it would be considerable work to carry out for a full ISA like RISC-V and
complex enforcement mechanism like the Depth Isolation micro-policy.
We therefore choose to validate the micro-policy of the previous section by
systematically \emph{testing} that it satisfies our properties. 
This focus on testing is better aligned with our immediate
goal of evaluating real enforcement mechanisms for real machine architectures.
Formal proof remains as possible future wrok.

We validate the micro-policy of the previous section by
systematically testing that it satisfies eager stack integrity and
confidentiality. We use a Coq specification of the RISC-V
architecture\footnote{Available at \url{https://github.com/mit-plv/riscv-coq}.}
%\leo{translated from Haskell via hs-2-coq?\apt{not worth saying since the MIT people did it}}
and extend it with a
runtime monitor implementing the stack-safety
micro-policy.\footnote{This code is {\em redacted for review}.\bcp{Why??}} We
  chose the Coq proof assistant as the setting for our implementation
  to ensure that our coinductive trace
  definitions are well-formed, and reason about them, and to leverage the power of the
  QuickChick property-based testing framework~\citep{Pierce:SF4}.

To use QuickChick, we build random test-case generators that produce
% \begin{enumerate}
% \item
(1) an initial RISC-V machine state, including most notably
  the program to be executed;
  % \item
(2)
  an initial policy state, tagging instructions corresponding to
  blessed call or return sequences appropriately, while marking
  all potential stack locations as $\tagNoDepth$; and
% \item
(3)
 a code map and a stack map \apt{what's that ??}, accurately characterizing the behavior
  of callers and callees.
% \end{enumerate}
%
That is, our generators need to produce not only a low-level program
that exhibits interesting control-flow patterns, but also the policy
meta-information and code annotations that would normally be produced
by the compiler.

@@

To write such generators we build on the work of
\citet{TestingNI:ICFP, DBLP:journals/jfp/HritcuLSADHPV16}, which
introduced {\em generation by execution} to produce progams that lead
machines towards interesting behaviors. Generation by
execution receives as an input a partially instantiated machine state
and attempts to generate an instruction (or a sequence of instructions
such as a blessed sequence) that makes sense locally (e.g., jumps go
to a potentially valid code location and loads read from a
potentially valid stack location). Then we step the machine and repeat
the process until we generate or execute some target number of
instructions, or reach a point where the machine can't step
anymore.

We extend this technique to keep track of the control flow behavior of
the program being generated: each time a call or return sequence is
generated, we ensure that the appropriate policy tags and code
annotations are set for the entry or return points. At the same time,
we allow the generation to sometimes relax those constraints,
introducing potentially ill-formed flows: this causes our programs
to failstop when executed in conjunction with the policy monitor,
but also allows for revealing errors in our setup.\apt{??}

@@

%The mean-time-to-failure (MTTF) for various bugs can be found in
%Table~\ref{tab:bug-table}. All experiments were run in a desktop
%machine equipped with i7-8700 CPU @ 3.20GHz with 32GB
%RAM.
% \leo{Something like: There is really no baseline to compare this
  %against, but include it here for the future/reasons?}
%
The artificially injected bugs weaken the tag propagation
and checking mechanism.
%
For example, when loading from a stack location, the correct
micro-policy needs to enforce that the tag on the location being read
is $\tagStackDepth{n}$ for some number $n$ and that the tag of the
current $\PCname$ is $\tagPCDepth{n}$ for the same depth $n$. We can relax
that restriction by not checking the depth equality (row {\em
  LOAD\_NO\_CHECK}).
%
Similarly, when storing to a stack location, the correct micro-policy
needs to ensure that the tag on the memory location is either
$\tagNoDepth$ or has again the same depth as the current $\PCname$
tag. Relaxing that constraint causes violations to the integrity
property (row {\em STORE\_NO\_CHECK}).

@@
